---
title: "Plots US economy"
author: "Philipp Hauber"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=6, #fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```


```{r}
# clear workspace
rm(list = ls())
# turn off scientific notation
options(scipen = 999)
# working directory
setwd("C:/Users/hauber/Documents/GitHub/plots-US-econ")
```

# Rig count and drilling activity

## Baker Hughes data

Data are from [Baker Hughes webpage](<https://rigcount.bakerhughes.com/na-rig-count/>) but need to be downloaded and stored in the repo. Currently, the xlsb-file and sheet with the historical data are called (*this should be the first thing to check if something goes wrong!*):

```{r, echo = TRUE}
filename <- "North America Rotary Rig Count (Jan 2000 - Current).xlsb"
sheetname <- "US Oil & Gas Split"
```

```{r}
library(lubridate)
library(dplyr)
library(tidyr)
library(readxlsb)
library(ggplot2)

# read in data
df_bh <- read_xlsb(path = paste0(getwd(), "/", filename),
                  sheet = sheetname,
                  col_names = TRUE,
                  skip = 6)

# overwrite dates
start_date <- "1987-07-17"
df_bh$Date <- seq(as.Date(start_date), length.out = nrow(df_bh), by = "1 week")

# convert from weekly to monthly frequency
df_bh %>% gather(variable, value_ww, -Date) %>%
          filter(!(variable %in% c("X..Gas", "X..Oil"))) %>%
          mutate(qq = ceiling(month(Date) / 3), yy = year(Date)) %>%
          group_by(yy, qq, variable) %>%
          summarise(value = round(mean(value_ww, na.rm = TRUE), digits = 0)) %>%
          ungroup() %>%
          mutate(date = make_date(year = yy, month = 3 * qq)) %>%
          select(date, variable, value, -yy, -qq) -> df_bh
```

Weekly dates aren't properly read in so we overwrite them manually with starting date `r start_date` (*periodically check that this is still the case!*) and convert to quarterly frequency. Furthermore, to plot the rig count along with drilling activity we need to rebase the series (2012 = 100):

```{r}

# calculate mean in 2012
tmp <- df_bh %>% mutate(yy = year(date)) %>%
                 filter(yy == 2012) %>%
                 group_by(yy, variable) %>%
                 summarise(value2012 = mean(value))

# merge with original data to calculate rebased values
df_bh_rebased <- merge(df_bh, tmp, by = "variable") %>%
                 mutate(value_rebased = 100 * value / value2012) %>%
                 select(date, variable, value = value_rebased, -value, -value2012, -yy) %>%
                 mutate(units = "2012 = 100")

# remove tmp dataframe
rm(tmp)

# combine with original data plus additional column indicating transformation
df_bh %>% mutate(units = "number") %>% rbind(df_bh_rebased) -> df_bh

# remove dataframe after merge
rm(df_bh_rebased)
```

## Federal Reserve industrial production data (G.17)

As a measure of drilling activity we use the subindex *oil and gas well drilling* from the Federal Reserve's Industrial Production (G.17) release. The series is sourced from FRED (**IPN213111S**)

```{r}
library(fredr)
fredr_set_key("84efd638e2db29ef758e0d8e081a4c05")

# download industrial production data
tmp_ip <- fredr(series_id = "IPN213111S", 
               observation_start = as.Date("1987-01-01"),
               frequency = "q"
              ) 

# download real PFI:mining, wells and shafts
tmp_inv <- fredr(series_id = "E318RX1Q020SBEA", 
               observation_start = as.Date("1987-01-01"),
               frequency = "q"
              ) 

# merge into one dataframe and relabel series_id column
df_fred <- rbind(tmp_ip, tmp_inv)
df_fred <- select(df_fred, date, variable = series_id, value)

# remove tmp dataframes
rm(tmp_ip, tmp_inv)
```

## Plot number of rigs vs drilling activity

Merge datasets and produce plot of total rigs (oil + gas + miscellaneous) and drilling activity

```{r}
# plot total number of rigs and drilling activity
df_bh %>% filter(units == "2012 = 100") %>% 
          select(-units) %>%
          rbind(df_ip) %>% 
          filter(date >= "1990-01-01", variable %in% c("Total", "IPN213111S")) %>% 
          ggplot(aes(x = date, y = value, col = variable))+
          geom_line()+
          ylab("Index (2012 = 100)")+
          xlab("")+
          scale_color_discrete(name = "", labels = c("ip: oil & gas well drilling", "rig count"))+
          scale_x_date(date_breaks = "4 years", date_labels = "%Y")+
          theme(axis.text.x = element_text(angle = 0))+
          labs(title = "United States: drilling activity",
               subtitle = "number of rotary rigs and industrial production",
               caption = "Quarterly data. Source: Baker Hughes, Federal Reserve, Bureau of Economic Analysis")
#ggsave("plot_rigs_ip.png")
```

Another interesting aspect of the Baker Hughes rig count data is the split between oil and gas wells:

```{r}
df_bh %>% filter(variable %in% c("Oil", "Gas"), 
                 date >= as.Date("1990-01-01"), 
                 units == "number") %>%
          ggplot(aes(x = date, y = value, col = variable))+
          geom_line()+
          scale_color_discrete(name = "", labels = c("gas", "oil"))+
          labs(title = "Number of rotary rigs in the United States",
               caption = "Monthly data. Source: Baker Hughes",
               y = "", x = "")
#ggsave("plot_oil_and_gas_rigs.png")
```

## Export data

```{r}
df_export <- df_bh %>% filter(units == "number", variable == "Total") %>%
                       select(-units) %>%
                       rbind(df_ip) %>%
                       filter(date >= as.Date("1990-01-01")) %>%
                       spread(variable, value) %>%
                       select(date, ip_drilling = IPN213111S, rigcount = Total)

#write.table(file = "rigcount_ip.csv", df_export, sep = ";", dec = ".", row.names = FALSE, na = "#NV")
```

